---
title: "Innovation Landscapes:  An Analysis of National Science Foundation Research Grants from 1999-2016"
author: "Sahar Alhassan, Kaitlyn Lynes, Deepa Mehta, Ilya Perepelitsa"
date: "December 21, 2015"
output: pdf_document
---
\begin{center}
\section{Introduction}
\end{center}
Innovation is often considered to be an engine for development. Governments typically support innovation by funding research and development for universities, corporations, and other institutions (Etzkowitz and Leydesdorff 2000). Recently, the White House approved a bipartisan budget agreement allocating higher funding over the next few years in order to maintain the United States' leadership in innovation (Donovan 2015). One of the few agencies that this bipartisan agreement will support is the National Science Foundation (NSF). 

NSF is an independent, federal grantmaking agency that awards nearly one fourth of all federal grants in the United States, nearly 11,000 new grants per year, and has a current annual budget of $7.3 billion (NSF 2015). A majority of these grants support critical research in science and technology. In order to get a sense of how innovation research, as procured by the NSF, is distributed, one can visit the NSF website to locate some summary statistics about grant awards. However, the NSF website provides only basic information about the number of grants awarded and grant allocations by states and institutions. For any serious researcher who is interested in understanding the landscape of federal funding of innovation in the United States, one is met with an aggregate dataset that numbers in the millions of grants awarded since NSF's founding in 1950. 

Given the sheer number of grants and the number of variables that denote grant type, this report began with a desire to develop an analytical methodology to understand the distribution of NSF grants and recognize any emergent patterns within these numbers. As such, this report studies NSF grants from the fiscal years 1999 to 2016 and aims to understand the distribution of NSF grants during this period. By developing a subset of the aggregate NSF data, we have been able to study the distribution of grant duration by type of grant (new grants, continuing grants, or revised grants), grants by recipient type (universities, private instititutions or corporations, etc.), by state, and by types of funded research. 

The broader aim of this exercise in big data analytics is to be able to apply quantitative methods to larger datasets, acquire subsets, and devise collated dataframes that can then be useful in identifying distributional attributes. The findings in this report and dataframe may be useful to scholars, grantseeking organizations, and individuals, as well as state agencies and private institutions to better understand the landscape of research funding in the United States. 
\begin{center}
\section{Methodology}
\end{center}
As a starting point, our dataset is comprised of queries obtained from the publicly available https://www.usaspending.gov. Using multiple queries on the following page https://www.usaspending.gov/Pages/TextView.aspx?data=HomeAwardTypeFunding , the range of fiscal years 1999-2016 was obtained. The variables that were selected for the working dataframe are as follows: 

#####Grant Periods
* fiscal_year - Budget year
* starting_date - Start date of awarded research project 
* ending_date - End date of awarded research project
* obligation_action_date - Start date of grant award obligation 

#####Grant Amounts and Grant Type
* fed_funding_amount - Dollar amount of grant 
This is the total amount awarded for a particular grant.

* action_type - Type of Grant Award  
This variable tells us whether this is a new grant, a continuing grant, or revised grant. A new grant refers to a new award which typically entail a "specific level of support for a specified period of time", while continuing grants provide specified support for an initial period of time, "with a statement of intent to provide additional support of the project for additional periods, provided funds are available", while revised grants refer to existing NSF grants that have been amended with new terms (NSF 2015). Revised grants differ from continued grants in that some aspect of a grant's funding needs appears to have changed, and these Revised grants must now compete direclty with new grants. 

#####Grant Recipient Information 
* recipient_name - Recipient Name  
This variable denotes the name of the recipient, usually a university or organization. 

* recipient_type -Type of Recipient  
This variable allows us to identify whether the recipeint organization is a higher education institution, corporation, or non-profit, among other categories.

* cfda_program_title - Field of research  
This refers to the research discipline of a grant. 

#####Grant Recipient Location 
* recipient_city_name - Recipient's City  
* principal_place_state - Recipient location state  
* principal_place_state_code - Recipient location state code
These variable allows us to identify the location of the recipient.  

The following modifications were conducted to the existing variables:
1. conversion of non-numeric variable into categorical variables   
2. dates were transformed to be treated as dates by R  
3. dublicates of variable values were merged to be treated as identcal values     
4. cfda_programm_title values were replaced by more concise Continue, New and Revision   

The following variables were added as new columns
* res_duration - This is the duration of research obtained by calculating the difference between the ending and starting dates of research. Such values were divided by 30 to obtain duration in months.

* costmon - The variable fed_funding_amount was divided by res_duration to establish the common base for comparing amounts alocated to different recipients.

* monstar, mondec - Month of research start and obligation action date respectively.

* logcost - Log-transormed costmon calculated for plotting these data.

The resulting dataset consists of 410,292 observations of 17 variables. Various subsets were created to calculate cumulative amounts granted to states, programs and by months. Subsets of top recipient states was also created for similar purposes. 

\begin{center}
\section{General observations}
\end{center}

Initially, we mapped the total number of awarded grants (410,292) from 1999 to 2016 based on the _Obligation Action Date_ and _Awarded Grant Dollars per month_ divided by _Action Type_. Here, we observed a concentration of continuing grants using a log transformation to identify any relationships between the type of grant over this 17 year period. We also observed a steady number of all grants to be somewhere between $0 to $500,000, with some continuing awards reaching $1,000,000 and some new awards reaching $1,500,000. We also observed revised grants in the negatives. One explanation for these negative numbers of nominal grant dollars (mostly in revision grants) can be that these grants often seek to make up for cost overruns. Another explanation is that negative allocations may also suggest that funding needs were lower than expected and thus the grantees may be obligated to return the unused funds. \vspace{20 mm}

```{r, fig.width=12, fig.height=4, echo=FALSE, warning=FALSE, error=FALSE}
load("NSF.RData")

newdata1 <- schoolsall[which(schoolsall$obligation_action_date < '2015-01-01' & schoolsall$obligation_action_date > '2012-01-01'), ]
newdata5 <- schoolsall[which(schoolsall$starting_date < '2015-01-01' & schoolsall$starting_date > '1999-01-01'), ]

bigstates <- schoolsall[which(schoolsall$principal_place_state_code == "CA" | schoolsall$principal_place_state_code == "CO" | schoolsall$principal_place_state_code == "DC" | schoolsall$principal_place_state_code == "IL"| schoolsall$principal_place_state_code == "MA"| schoolsall$principal_place_state_code == "NY"| schoolsall$principal_place_state_code == "PA"| schoolsall$principal_place_state_code == "TX"), ]

sumbystate <- aggregate(fed_funding_amount ~ principal_place_state_code + fiscal_year, schoolsall, sum)
sumbyprog <- aggregate(fed_funding_amount ~ cfda_program_title + fiscal_year, schoolsall, sum)
sumbymon <- aggregate(fed_funding_amount ~ mondec + fiscal_year, schoolsall, sum)
sumbyrect <- aggregate(fed_funding_amount ~ recipient_type + fiscal_year, schoolsall, sum)

library(ggplot2)

ggplot(schoolsall,aes(x=obligation_action_date, y = logcost)) + geom_point(aes(color= action_type ))  + ggtitle("log-transformed Grant Dollars per month of research, historical 1999-2016")+ labs(x="Obligation action date",y="log(Grant Dollars per month of research)") 

```

```{r, fig.width=12, fig.height=4, echo=FALSE, warning=FALSE, error=FALSE}
ggplot(schoolsall,aes(x=obligation_action_date, y = fed_funding_amount/100000)) + geom_point(aes(color= action_type )) + ggtitle("Nominal Grant Dollars per month of research / $100K, historical 1999-2016")+ labs(x="Obligation action date",y="Dollars per month of research / $100K") 

```
We then categorically analyzed the difference between continuing, new, and revised grants over the 17 year period using a log transformation of their allocated amounts. Here, we continued to see the steady flow of continuing grants, with some outliers of low grant awards. We also observed a more narrow, but steady concentration, of new grants, with even more outliers of lower grant awards. 

We also noted some seasonal cyclicity over the 16 year period, where approximatley every three years, there appears to be an upswing in grant amounts. \vspace{20 mm}

```{r, fig.width=12, fig.height=4, echo=FALSE, warning=FALSE, error=FALSE}
ggplot(newdata1, aes(y=logcost, x=obligation_action_date)) + geom_point(aes(color = action_type)) + facet_wrap(~action_type) + ggtitle("Seasonal cyclicality 2012-2015") + labs(x="Obligation action date, 2012-2015",y="log(Grant Dollars per month of research)") + 
  theme(
  axis.text.x = element_blank(),
  axis.text.y = element_blank(),
  axis.ticks = element_blank())
 
```
With this log transformation, we observed a large gap in the amount of revised grants. Analyzed in nominal grant allocation per month, we saw a steady flow of revised grants in perceived gap years in the middle of this 17 year cycle, with more grant amount fluctuations in other years. \vspace{20 mm}

```{r, fig.width=12, fig.height=3, echo=FALSE, warning=FALSE, error=FALSE}
ggplot(newdata5, aes(y=logcost, x=obligation_action_date)) + geom_point(aes(color = action_type))+  facet_wrap(~action_type) + ggtitle("A gap in revisions") + labs(x="Obligation action date, 1999-2015",y="log$") + 
  theme(
  axis.text.x = element_blank(),
  axis.text.y = element_blank(),
  axis.ticks = element_blank())
 
```

The plot of revisions shows a significant gap which might be explained by either no funding activity (no revisions in these years) or technicalities of log conversions - there were +inf and -inf values that had to be substituted by NA values and these values might have been concentrated in the time period reflected on this plot.\vspace{10 mm}

\vspace{10 mm}

```{r, fig.width=12, fig.height=2, echo=FALSE, warning=FALSE, error=FALSE}
ggplot(newdata5, aes(y=fed_funding_amount/100000, x=obligation_action_date)) + geom_point(aes(color = action_type))+  facet_wrap(~action_type) + ggtitle("Gap explained by log conversion") + labs(y="Nominal M/$100K",x=element_blank()) + 
  theme(
  axis.text.x = element_blank(),
  axis.text.y = element_blank(),
  axis.ticks = element_blank())
 
```
Plotting the nominal scale (not log-transformed) shows that the values in the gap years are present and are negative and that explains the negative infinite values. This adds a limitation of this research - decreasing the difficulty of producing visual insights  affects the quality of data. In this case it significantly decreases the qyality of data and can lead to wrong conclusions.  \vspace{20 mm}

In order to better understand the cost differentials of these awarded grants and to better understand the perceived gaps, we studied the density of the three action types over the 17 year period. The density plot reveals several factors about the distribution of our data points. For instance, we observe two steady peaks in the revised grants, while sharp peaks for both Continued and New grants. \vspace{20 mm}

```{r, fig.width=4, fig.height=2, echo=FALSE, warning=FALSE, error=FALSE}
ggplot(schoolsall, aes(x=log(costmon), fill=action_type)) + geom_density(alpha = 0.6) + ggtitle("Cost differences between action types") + labs(y="Density",x="log(Grant $/Month)")
```
We observed a strong bimodal cost trend for revised grants, both modes are lower than Continuing and New grants. New grants are generally larger and their frequency is higher than among Continuing grants. 

\vspace{20 mm}
```{r, fig.width=12, fig.height=5, echo=FALSE, warning=FALSE, error=FALSE}
ggplot(schoolsall, aes(y=res_duration, x=log(costmon))) + geom_point(aes(color = action_type))+ facet_wrap(~action_type)+ ggtitle("Difference in cost and duration")+ labs(y="Duration in Months",x="log(Grant $/Month)")
```
As for duration of grants, New grants do not last longer than 100 months, while Revised and Continuing grants are longer in duration. 

```{r, fig.width=15, fig.height=6, echo=FALSE, warning=FALSE, error=FALSE}
ggplot(schoolsall, aes(x=monstar, y=log(costmon), fill=action_type)) + geom_boxplot()+ facet_wrap(~action_type) + ggtitle("Monthly Outliers") + theme(axis.text.x = element_text(angle=45)) + labs(y="log(Grant $/Month)",x=element_blank())
```
We can also mark a decrease in variance towards the end of the year. The greatest number of outliers are for New and Continuing grants in summer and fall months. The greatest number of positive outliers in are visible in Revised grants for the same duration periods as the other types.

\begin{center}
\section{Fields of Research}
\end{center}

True to the NSF's mission to advance science and technology, STEM fields consistently receive the highest annual total allocations for each grant type, receiving approximately $500 million to $1.5 billion annually. There is a genereal increase in Education and Human Resources grants. Generally, the distributions among all fields of research appears to be relatively normal with small skews, while variances are very high. 


```{r, fig.width=15, fig.height=6, echo=FALSE, warning=FALSE, error=FALSE}
ggplot(sumbyprog, aes(x=fed_funding_amount, y=reorder(cfda_program_title, fed_funding_amount), fill = fiscal_year)) + ggtitle("Most financed fields of research") + labs(y=element_blank(),x="Cumulative annual allocation") + geom_point(size=3) + theme(axis.text.x = element_text(angle=60, hjust = 1), panel.grid.major.y = element_blank(), panel.grid.minor.y = element_blank(), panel.grid.major.x = element_line(color="grey60", linetype='dashed'))
```

\begin{center}
\section{Grant Receiving States}
\end{center}

California, New York, Massachussetts and DC are the states with the highest annual cimilative allocations. Annual variance is high for these four but being a recipient state with the most allocations is consistent across 1999-2016. 

The range is significant - each tick in the plot below represents $300 million. For example, California's grant allocations range from $500 million to $1 billion, which represents the difference of $500 million between the highest and lowest years for this state.

```{r, fig.width=15, fig.height=10, echo=FALSE, warning=FALSE, error=FALSE}
ggplot(sumbystate, aes(x=fed_funding_amount, y=reorder(principal_place_state_code, fed_funding_amount), fill = fiscal_year)) + ggtitle("Most financed states") + labs(y=element_blank(),x="Cumulative annual allocation") +
  geom_point(size=3) + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle=60, hjust = 1),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.x = element_line(color="grey60", linetype='dashed'))
```

```{r, fig.width=15, fig.height=6, echo=FALSE, warning=FALSE, error=FALSE}
ggplot(bigstates, aes(x=principal_place_state_code, y=log(costmon), fill = action_type)) + geom_boxplot() + facet_wrap(~action_type) + guides(colour = guide_legend(override.aes = list(size=12)))+ ggtitle("Most financed states - outliers") + labs(x=element_blank(),y="log(Grant $/Month)")
```

California, New York and DC consistently show more negative otliers (small grants) than other states given that medians with minor variations are roughly similar for most of the larfest states and the upper bounds for larger grants are roughly the same as well.

\begin{center}
\section{Frequency of Grant Allocations by Month}
\end{center}

```{r, fig.width=15, fig.height=6, echo=FALSE, warning=FALSE, error=FALSE}
ggplot(sumbymon, aes(x=fed_funding_amount, y=mondec, fill = fiscal_year)) + ggtitle("Highest cumulative months values") +
  geom_point(size=3) + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle=60, hjust = 1),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.x = element_line(color="grey60", linetype='dashed'))

```
The largest grants are consistently allocated in the summer months as well as September, while grant activity significantly slows down in the fall and winter months. Grant activity tends to peak in August and is slightly lower in September. From this data, we can presuppose that the general annual NSF cycle for grant awards appears to be from October to September. However, start dates and end dates for research may vary. 

We also observe month to month shifts in frequency, most significantly for Revised grants, generating a bimodal distribution in this regard. Continuing and New grants show some shifts in allocations; their frequence also varies over time, making frequency values more or less distinct depending on the month. 

The final plot here shows that there also appears to be a consistent increase in variance during the summer, after which grant allocation activity generally slows down in September. This plot also demonstrates a strong positive skew with many negative outliers. 

```{r, fig.width=15, fig.height=6, echo=FALSE, warning=FALSE, error=FALSE}

ggplot(schoolsall, aes(x=log(costmon), fill=mondec)) + geom_density(alpha = 0.1) + facet_wrap(~action_type) +geom_vline(aes(xintercept=mean(logcost), na.rm=T), color="red", linetype="dashed", size=1) + ggtitle("Monthly variations") + labs(x="log(Grant $/Month)",y="Density")
```

```{r, fig.width=15, fig.height=6, echo=FALSE, warning=FALSE, error=FALSE}
ggplot(schoolsall, aes(x=mondec, y=log(costmon), fill=mondec)) + geom_boxplot() + facet_wrap(~action_type)+ ggtitle("Fall variance") + labs(x=element_blank(),y="log(Grant $/Month)")+ theme(axis.text.x = element_text(angle=45))

```
\begin{center}
\section{Grant Recipients}
\end{center}

The highest cumulative amounts is found amongst state and private higher education institutions and "All other". However, the dispersal amongst them is also quite high. The data appearss to show that there is no fixed proportion of total grant dollars that higher education institutions are receiving each year. 

```{r, fig.width=15, fig.height=6, echo=FALSE, warning=FALSE, error=FALSE}
ggplot(sumbyrect, aes(x=fed_funding_amount, y=reorder(recipient_type, fed_funding_amount), fill = fiscal_year)) + ggtitle("Concentrations by types of recipients - cumulative annual") + 
  geom_point(size=3) + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle=60, hjust = 1),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.x = element_line(color="grey60", linetype='dashed'))

```

\begin{center}
\section{Discussion}
\end{center}



\begin{center}
\section{Bibliography}
\end{center}

Donovan, Shaun. “A Critical Step Forward for a Stronger Economy and a Simpler, Fairer Tax Code.” 2015. Whitehouse.gov. December 18. https://www.whitehouse.gov/blog/2015/12/18/critical-step-forward-stronger-economy-and-simpler-fairer-tax-code.

Etzkowitz, Henry, and Loet Leydesdorff. 2000. “The Dynamics of Innovation: From National Systems and ‘Mode 2’ to a Triple Helix of University–industry–government Relations.” Research Policy 29 (2). Elsevier: 109–23.

“NSF - National Science Foundation.” 2015. Accessed December 21. Retrieved from http://www.nsf.gov/.
